# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r}
```

I'm assuming that the current working directory has been set to the directory where the
data is -- this could be the top-level directory in the Git repository. I'm first unzipping,
which produces a .csv file, and then reading the csv file. I also print the summary to show
that the data has actually been read and to give an idea of what it's like.

```{r}
unzip('activity.zip')
data <- read.csv('activity.csv')
summary(data)
```

### Preprocessing

We will need a dataset that has the days (date) and the total steps for that day, one row per day. I store this in the stepsdata data frame.


```{r}
totalsteps <- tapply(data$steps, data$date, FUN="sum")
stepsdata <- as.data.frame(cbind(names(totalsteps), totalsteps))
```



## What is mean total number of steps taken per day?

The code here produces a histogram of the mean of the total number of steps taken each day.
Each day has 288 data points, corresponding to one point every 5 minutes.

```{r}
hist(totalsteps, xlab="Number of steps per day", main = NULL)
```

Calculate the mean number of steps per day:

```{r}
mean(totalsteps, na.rm=TRUE)
```

Calculate the median number of steps per day:

```{r}
median(totalsteps, na.rm=TRUE)
```




## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
